Clinical prediction with localized modeling using similarity-based cohorts: A scoping review

### 2023-09-10 docs/Manuscript/Manuscript.Rmd
make additional edits during resubmission process

We discuss in this review a hybrid computational approach we call _localized modeling_ (see Section \ref{sec:coherence}) that appears to have been introduced independently several times. The approach fits ML models, including regularized families of generalized regression models (GLMs), to cohorts retrieved via similarity matching. Whereas CBR and NN prediction use patient similarity to obtain an ideally homogeneous subpopulation matched to the index case, in localized modeling one seeks a relevant yet heterogeneous subpopulation, each member of which may deviate from the index case.
A prediction or inference is then obtained from a model fitted to this _similarity cohort_.
The use of similarity cohorts distinguishes localized modeling from classical GLMs that still dominate analysis in clinical trials and comparative effectiveness research, as well as from many cutting-edge ML models that are fitted to whole populations, while the cohorts' heterogeneity and its utility to the model distinguish the approach from CBR.
In principle, this allows interpretable models like GLMs to identify and measure the effects of determining factors that are specific to the index case.
In the reviewed studies, localized modeling shows some promise toward simultaneously achieving the performance improvements of ML and the interpretability of classical models.

@@

Here we describe our review process, including deviations from plans and the reasons for them.
More details are included in Section \ref{sec:appendix-selection}.

@@

We derived the eligibility criteria itemized in the Introduction from a seed set of previously read studies. Based on these criteria, we formulated search strings for five platforms: PubMed, Web of Science, Academic Search Premier, Google Scholar, and MathSciNet. We finalized the PubMed search first, then adapted it to the other platforms (Section \ref{sec:appendix-search}).

@@

## Full text review
\label{sec:full-text}

@@

Following the selection sources aligning with our review, we collected characteristics of included studies (Section \ref{sec:data}). The features included bibliographic fields (date of publication, journal, authors, title, keywords, DOI), study goals (objective, generalizable knowledge, evaluation, clinical/medical domain), data sets (data source, type of data, range of data, number of cases/incidences, number of predictors/features), and methodological choices (types of similarity measure, families of adaptation step/statistical model, method(s) compared against, performance measures, results of evaluations and comparisons, name given to modeling approach). We used these data to detect and visualize trends amongst the included studies.

@@

```{r fig:prisma, fig.align='center', out.width="80%", echo=FALSE}
#| fig.cap =
#| "\\label{fig:prisma}
#| PRISMA-S flow chart.
#|  Lowercase letters refer to items obtained from
#|  the seed set (m),
#|  the structured search (n), and
#|  citation/reference tracking (s)."
knitr::include_graphics("../../fig/fig-prisma.pdf")
```

@@

The `r nrow(tab_synthesis)` included studies were analyzed based on several characteristics (see Section \ref{sec:data}), and we report and describe some observations here (Table \ref{tab:synthesis}). The studies were published in a variety of journals, with some of greater frequency, though no single journal published more than 3. The journal _Evolving Systems_ published 2 of the included studies, _Artificial Intelligence in Medicine_ published 3, _Hindawi - Journal of Healthcare Engineering_ published 3, and _PLOS One_ published 2.

@@

```{r fig:year, fig.align='center', out.width="75%", echo=FALSE}
#| fig.cap =
#| "\\label{fig:year}
#| Number of publications in our sample each year."
knitr::include_graphics("../../fig/fig-years.png")
```

@@

```{r fig:properties, fig.align='center', out.width="100%", echo=FALSE}
#| fig.cap =
#| "\\label{fig:properties}
#| Share of studies characterized by several design elements:
#|  data source (row),
#|  clinical task (fill color), and
#|  remit (opacity)."
knitr::include_graphics("../../fig/fig-properties.png")
```

## Rationales
\label{sec:rationales}

The included studies hypothesized, asserted, or assumed several benefits of localized modeling specific to clinical and medical settings and advantages over other modeling approaches (see Section \ref{sec:data}).
The most common was that the restriction to similar or relevant past cases would improve predictive performance for the index case [@Mariuzzi1997; @Liang2015a; @Ng2015; @Lee2017]. In particular, @Lee2015 hypothesized and confirmed that the value of each past case would be positively related to its similarity to the index case, an assumption built in to the weighting schemes of other approaches. @Lowsky2013 made a different case, that the fewer parametric assumptions and complications of a CBR-style model would allow for greater accuracy.
Additionally, several investigators asserted that the use of localized cohorts befit the clinical focus on the individual patient rather than the population, without reference to performance [@Song2006c; @Xu2008a].

@@

```{r fig:methods, fig.align='center', out.width="100%", echo=FALSE}
#| fig.cap =
#| "\\label{fig:methods}
#| Frequency of recurring methodological elements across included studies."
knitr::include_graphics("../../fig/fig-methods.png")
```

## Performance assessments
\label{sec:performance}

@@

## Identified needs
\label{sec:needs}

@@

## Coherence
\label{sec:coherence}

@@

```{r fig:framework, fig.align='center', out.width="80%", echo=FALSE}
#| fig.cap =
#| "\\label{fig:framework}
#| General framework for localized models."
knitr::include_graphics("../../fig/fig-typology.pdf")
```

@@

As discussed in Section \ref{sec:rationales}, most studies were premised on the potential predictive value of localized models.
As cautioned in Section \ref{sec:performance}, not all experiments affirmed this premise. Most of those that did involved using metric learning to improve retrieval and sometimes iterative optimization of the metric and of the localized model.
Some studies showcased results using a variety of such specializations [@CampilloGimenez2012a; @Ng2015; @Zhang2018a]. However, because these advanced implementations were mostly compared against their precursors or commonplace alternatives, we cannot speak to their performance or any trade-offs with respect to each other.
This would be a worthy goal of future work.

While most studies identified outstanding technical needs (Section \ref{sec:needs}), few emphasized the need to assess human-focused qualities like user interface and user experience, interpretability, meaningfulness, or trust.
Most studies identified interpretability as an advantage of localized models, though none evaluated model interpretability and few proposed new interpretative uses.
Lack of trust in ML tools is a long-recognized problem that direct intepretability of model components could help alleviate, but this is more often assumed than demonstrated.
One valuable direction for future work would be to measure the utility of interpretable components in research and in clinical practice and the correctness and confidence of users in their interpretations.

@@

# Data and code
\label{sec:data}

@@

## Purpose of review
\label{sec:appendix-purpose}

@@

## Search
\label{sec:appendix-search}

@@

## Screening process
\label{sec:appendix-screening}

@@

## Selection process
\label{sec:appendix-selection}

Some PDFs were obtained using Zotero from a university workstation, and the remaining were obtained through university library services. The review process is detailed in Section \ref{sec:full-text}.

>>>>

We discuss in this review a hybrid computational approach we call _localized modeling_ (see Section \ref{coherence}) that appears to have been introduced independently several times. The approach fits ML models, including regularized families of generalized regression models (GLMs), to cohorts retrieved via similarity matching. Whereas CBR and NN prediction use patient similarity to obtain an ideally homogeneous subpopulation matched to the index case, in localized modeling one seeks a relevant yet heterogeneous subpopulation, each member of which may deviate from the index case.
A prediction or inference is then obtained from a model fitted to this _similarity cohort_.
The use of similarity cohorts distinguishes localized modeling from classical GLMs that still dominate analysis in clinical trials and comparative effectiveness research, as well as from many cutting-edge ML models that are fitted to whole populations, while the cohorts' heterogeneity and its utility to the model distinguish the approach from CBR.
In principle, this allows interpretable models like GLMs to identify and measure the effects of determining factors that are specific to the index case.
In the reviewed studies, localized modeling shows some promise toward simultaneously achieving the performance improvements of ML and the interpretability of classical models.

@@

Here we describe our review process, including deviations from plans and the reasons for them.
More details are included in Section \ref{selection-process}.

@@

We derived the eligibility criteria itemized in the Introduction from a seed set of previously read studies. Based on these criteria, we formulated search strings for five platforms: PubMed, Web of Science, Academic Search Premier, Google Scholar, and MathSciNet. We finalized the PubMed search first, then adapted it to the other platforms (Section \ref{search-design}).

@@

## Full text review

@@

Following the selection sources aligning with our review, we collected characteristics of included studies (Section \ref{data-and-code}). The features included bibliographic fields (date of publication, journal, authors, title, keywords, DOI), study goals (objective, generalizable knowledge, evaluation, clinical/medical domain), data sets (data source, type of data, range of data, number of cases/incidences, number of predictors/features), and methodological choices (types of similarity measure, families of adaptation step/statistical model, method(s) compared against, performance measures, results of evaluations and comparisons, name given to modeling approach). We used these data to detect and visualize trends amongst the included studies.

@@

```{r prisma, fig.align='center', out.width="80%", echo=FALSE}
#| fig.cap =
#| "PRISMA-S flow chart.
#|  Lowercase letters refer to items obtained from
#|  the seed set (m),
#|  the structured search (n), and
#|  citation/reference tracking (s)."
knitr::include_graphics("../../fig/fig-prisma.pdf")
```

@@

The `r nrow(tab_synthesis)` included studies were analyzed based on several characteristics (see Section \ref{data-and-code}), and we report and describe some observations here (Table \ref{tab:synthesis}). The studies were published in a variety of journals, with some of greater frequency, though no single journal published more than 3. The journal _Evolving Systems_ published 2 of the included studies, _Artificial Intelligence in Medicine_ published 3, _Hindawi - Journal of Healthcare Engineering_ published 3, and _PLOS One_ published 2.

@@

```{r year, fig.align='center', out.width="75%", echo=FALSE}
#| fig.cap =
#| "Number of publications in our sample each year."
knitr::include_graphics("../../fig/fig-years.png")
```

@@

```{r properties, fig.align='center', out.width="100%", echo=FALSE}
#| fig.cap =
#| "Share of studies characterized by several design elements:
#|  data source (row),
#|  clinical task (fill color), and
#|  remit (opacity)."
knitr::include_graphics("../../fig/fig-properties.png")
```

## Rationales
\label{rationales}

The included studies hypothesized, asserted, or assumed several benefits of localized modeling specific to clinical and medical settings and advantages over other modeling approaches (see Section \ref{data-and-code}).
The most common was that the restriction to similar or relevant past cases would improve predictive performance for the index case [@Mariuzzi1997; @Liang2015a; @Ng2015; @Lee2017]. In particular, @Lee2015 hypothesized and confirmed that the value of each past case would be positively related to its similarity to the index case, an assumption built in to the weighting schemes of other approaches. @Lowsky2013 made a different case, that the fewer parametric assumptions and complications of a CBR-style model would allow for greater accuracy.
Additionally, several investigators asserted that the use of localized cohorts befit the clinical focus on the individual patient rather than the population, without reference to performance [@Song2006c; @Xu2008a].

@@

```{r methods, fig.align='center', out.width="100%", echo=FALSE}
#| fig.cap =
#| "Frequency of recurring methodological elements across included studies."
knitr::include_graphics("../../fig/fig-methods.png")
```

## Performance assessments
\label{performance-assessments}

@@

## Identified needs

@@

## Coherence

@@

```{r framework, fig.align='center', out.width="80%", echo=FALSE}
#| fig.cap =
#| "General framework for localized models."
knitr::include_graphics("../../fig/fig-typology.pdf")
```

@@

As discussed in Section \ref{rationales}, most studies were premised on the potential predictive value of localized models.
As cautioned in Section \ref{performance-assessments}, not all experiments affirmed this premise. Most of those that did involved using metric learning to improve retrieval and sometimes iterative optimization of the metric and of the localized model.
Some studies showcased results using a variety of such specializations [@CampilloGimenez2012a; @Ng2015; @Zhang2018a]. However, because these advanced implementations were mostly compared against their precursors or commonplace alternatives, we cannot speak to their performance or any trade-offs with respect to each other.
This would be a worthy goal of future work.

While most studies identified outstanding technical needs (Section \ref{identified-needs}), few emphasized the need to assess human-focused qualities like user interface and user experience, interpretability, meaningfulness, or trust.
Most studies identified interpretability as an advantage of localized models, though none evaluated model interpretability and few proposed new interpretative uses.
Lack of trust in ML tools is a long-recognized problem that direct intepretability of model components could help alleviate, but this is more often assumed than demonstrated.
One valuable direction for future work would be to measure the utility of interpretable components in research and in clinical practice and the correctness and confidence of users in their interpretations.

@@

# Data and code

@@

## Purpose of review

@@

## Search design

@@

## Screening process

@@

## Selection process

Some PDFs were obtained using Zotero from a university workstation, and the remaining were obtained through university library services. The review process is detailed in Section \ref{full-text-review}.

### 2023-09-10 docs/Manuscript/Manuscript.tex
make additional edits during resubmission process

We discuss in this review a hybrid computational approach we call
\emph{localized modeling} (see Section \ref{sec:coherence}) that appears
to have been introduced independently several times. The approach fits
ML models, including regularized families of generalized regression
models (GLMs), to cohorts retrieved via similarity matching. Whereas CBR
and NN prediction use patient similarity to obtain an ideally
homogeneous subpopulation matched to the index case, in localized
modeling one seeks a relevant yet heterogeneous subpopulation, each
member of which may deviate from the index case. A prediction or
inference is then obtained from a model fitted to this \emph{similarity
cohort}. The use of similarity cohorts distinguishes localized modeling
from classical GLMs that still dominate analysis in clinical trials and
comparative effectiveness research, as well as from many cutting-edge ML
models that are fitted to whole populations, while the cohorts'
heterogeneity and its utility to the model distinguish the approach from
CBR. In principle, this allows interpretable models like GLMs to
identify and measure the effects of determining factors that are
specific to the index case. In the reviewed studies, localized modeling
shows some promise toward simultaneously achieving the performance
improvements of ML and the interpretability of classical models.

@@

\label{sec:full-text}

@@

\caption{\label{fig:prisma}PRISMA-S flow chart. Lowercase letters refer to items obtained from the seed set (m), the structured search (n), and citation/reference tracking (s).}\label{fig:fig:prisma}
\end{figure}

@@

The 25 included studies were analyzed based on several characteristics
(see Section \ref{sec:data}), and we report and describe some
observations here (Table \ref{tab:synthesis}). The studies were
published in a variety of journals, with some of greater frequency,
though no single journal published more than 3. The journal
\emph{Evolving Systems} published 2 of the included studies,
\emph{Artificial Intelligence in Medicine} published 3, \emph{Hindawi -
Journal of Healthcare Engineering} published 3, and \emph{PLOS One}
published 2.

@@

\caption{\label{fig:year}Number of publications in our sample each year.}\label{fig:fig:year}
\end{figure}

@@

\caption{\label{fig:properties}Share of studies characterized by several design elements: data source (row), clinical task (fill color), and remit (opacity).}\label{fig:fig:properties}
\end{figure}

@@

\label{sec:rationales}

The included studies hypothesized, asserted, or assumed several benefits
of localized modeling specific to clinical and medical settings and
advantages over other modeling approaches (see Section \ref{sec:data}).
The most common was that the restriction to similar or relevant past
cases would improve predictive performance for the index case
{[}11,15,16,26{]}. In particular, {[}14{]} hypothesized and confirmed
that the value of each past case would be positively related to its
similarity to the index case, an assumption built in to the weighting
schemes of other approaches. {[}12{]} made a different case, that the
fewer parametric assumptions and complications of a CBR-style model
would allow for greater accuracy. Additionally, several investigators
asserted that the use of localized cohorts befit the clinical focus on
the individual patient rather than the population, without reference to
performance {[}21,23{]}.

@@

\caption{\label{fig:methods}Frequency of recurring methodological elements across included studies.}\label{fig:fig:methods}
\end{figure}

@@

\label{sec:performance}

@@

\label{sec:needs}

@@

\label{sec:coherence}

@@

\caption{\label{fig:framework}General framework for localized models.}\label{fig:fig:framework}
\end{figure}

@@

As discussed in Section \ref{sec:rationales}, most studies were premised
on the potential predictive value of localized models. As cautioned in
Section \ref{sec:performance}, not all experiments affirmed this
premise. Most of those that did involved using metric learning to
improve retrieval and sometimes iterative optimization of the metric and
of the localized model. Some studies showcased results using a variety
of such specializations {[}15,27,30{]}. However, because these advanced
implementations were mostly compared against their precursors or
commonplace alternatives, we cannot speak to their performance or any
trade-offs with respect to each other. This would be a worthy goal of
future work.

While most studies identified outstanding technical needs (Section
\ref{sec:needs}), few emphasized the need to assess human-focused
qualities like user interface and user experience, interpretability,
meaningfulness, or trust. Most studies identified interpretability as an
advantage of localized models, though none evaluated model
interpretability and few proposed new interpretative uses. Lack of trust
in ML tools is a long-recognized problem that direct intepretability of
model components could help alleviate, but this is more often assumed
than demonstrated. One valuable direction for future work would be to
measure the utility of interpretable components in research and in
clinical practice and the correctness and confidence of users in their
interpretations.

@@

\label{sec:data}

@@

\label{sec:appendix-purpose}

@@

\hypertarget{search-1}{%
\subsection{Search}\label{search-1}}

\label{sec:appendix-search}

@@

\label{sec:appendix-screening}

@@

\label{sec:appendix-selection}

Some PDFs were obtained using Zotero from a university workstation, and
the remaining were obtained through university library services. The
review process is detailed in Section \ref{sec:full-text}.

>>>>

We discuss in this review a hybrid computational approach we call
\emph{localized modeling} (see Section \ref{coherence}) that appears to
have been introduced independently several times. The approach fits ML
models, including regularized families of generalized regression models
(GLMs), to cohorts retrieved via similarity matching. Whereas CBR and NN
prediction use patient similarity to obtain an ideally homogeneous
subpopulation matched to the index case, in localized modeling one seeks
a relevant yet heterogeneous subpopulation, each member of which may
deviate from the index case. A prediction or inference is then obtained
from a model fitted to this \emph{similarity cohort}. The use of
similarity cohorts distinguishes localized modeling from classical GLMs
that still dominate analysis in clinical trials and comparative
effectiveness research, as well as from many cutting-edge ML models that
are fitted to whole populations, while the cohorts' heterogeneity and
its utility to the model distinguish the approach from CBR. In
principle, this allows interpretable models like GLMs to identify and
measure the effects of determining factors that are specific to the
index case. In the reviewed studies, localized modeling shows some
promise toward simultaneously achieving the performance improvements of
ML and the interpretability of classical models.

@@

@@

\caption{PRISMA-S flow chart. Lowercase letters refer to items obtained from the seed set (m), the structured search (n), and citation/reference tracking (s).}\label{fig:prisma}
\end{figure}

@@

The 25 included studies were analyzed based on several characteristics
(see Section \ref{data-and-code}), and we report and describe some
observations here (Table \ref{tab:synthesis}). The studies were
published in a variety of journals, with some of greater frequency,
though no single journal published more than 3. The journal
\emph{Evolving Systems} published 2 of the included studies,
\emph{Artificial Intelligence in Medicine} published 3, \emph{Hindawi -
Journal of Healthcare Engineering} published 3, and \emph{PLOS One}
published 2.

@@

\caption{Number of publications in our sample each year.}\label{fig:year}
\end{figure}

@@

\caption{Share of studies characterized by several design elements: data source (row), clinical task (fill color), and remit (opacity).}\label{fig:properties}
\end{figure}

@@

\label{rationales}

The included studies hypothesized, asserted, or assumed several benefits
of localized modeling specific to clinical and medical settings and
advantages over other modeling approaches (see Section
\ref{data-and-code}). The most common was that the restriction to
similar or relevant past cases would improve predictive performance for
the index case {[}11,15,16,26{]}. In particular, {[}14{]} hypothesized
and confirmed that the value of each past case would be positively
related to its similarity to the index case, an assumption built in to
the weighting schemes of other approaches. {[}12{]} made a different
case, that the fewer parametric assumptions and complications of a
CBR-style model would allow for greater accuracy. Additionally, several
investigators asserted that the use of localized cohorts befit the
clinical focus on the individual patient rather than the population,
without reference to performance {[}21,23{]}.

@@

\caption{Frequency of recurring methodological elements across included studies.}\label{fig:methods}
\end{figure}

@@

\label{performance-assessments}

@@

@@

@@

\caption{General framework for localized models.}\label{fig:framework}
\end{figure}

@@

As discussed in Section \ref{rationales}, most studies were premised on
the potential predictive value of localized models. As cautioned in
Section \ref{performance-assessments}, not all experiments affirmed this
premise. Most of those that did involved using metric learning to
improve retrieval and sometimes iterative optimization of the metric and
of the localized model. Some studies showcased results using a variety
of such specializations {[}15,27,30{]}. However, because these advanced
implementations were mostly compared against their precursors or
commonplace alternatives, we cannot speak to their performance or any
trade-offs with respect to each other. This would be a worthy goal of
future work.

While most studies identified outstanding technical needs (Section
\ref{identified-needs}), few emphasized the need to assess human-focused
qualities like user interface and user experience, interpretability,
meaningfulness, or trust. Most studies identified interpretability as an
advantage of localized models, though none evaluated model
interpretability and few proposed new interpretative uses. Lack of trust
in ML tools is a long-recognized problem that direct intepretability of
model components could help alleviate, but this is more often assumed
than demonstrated. One valuable direction for future work would be to
measure the utility of interpretable components in research and in
clinical practice and the correctness and confidence of users in their
interpretations.

@@

@@

@@

\hypertarget{search-design}{%
\subsection{Search design}\label{search-design}}

@@

@@

Some PDFs were obtained using Zotero from a university workstation, and
the remaining were obtained through university library services. The
review process is detailed in Section \ref{full-text-review}.
